---
title: "infected.pilot.analysis"
author: "Joy O'Brien"
date: "2026-08-31"
output: html_document
---

0. Set wd and figure theme
```{r}
setwd("~/GitHub/sporeAI")
```

## Figure theme
```{r}
library(paletteer)
library(ggplot2)
theme_set(theme_classic() +
  theme(axis.title = element_text(size = 16),
        axis.title.x = element_text(margin = margin(t = 15, b = 15)),
        axis.title.y = element_text(margin = margin(l = 15, r = 15)),
        axis.text = element_text(size = 13),
        axis.text.x = element_text(margin = margin(t = 5)),
        axis.text.y = element_text(margin = margin(r = 5)),
        #axis.line.x = element_line(linewidth = 1),
        #axis.line.y = element_line(linewidth = 1),
        axis.ticks.y.right = element_line(linewidth = 1),
        axis.text.y.right = element_text(size = 13, margin = margin(l = 5)),
        axis.line.y.right = element_line(linewidth = 0.1),
        axis.ticks.x.top = element_line(linewidth = 1),
        axis.text.x.top = element_text(size = 13, margin = margin(b = 5)),
        axis.line.x.top = element_line(linewidth = 0.1),
        axis.line.x = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.x = element_line(linewidth = 1),
        axis.ticks.y = element_line(linewidth = 1),
        axis.ticks.length = unit(.1, "in"),
        panel.border = element_rect(color = "grey20", fill = NA, linewidth = 1.5),
        legend.text = element_text(size = 14),
        strip.text = element_text(size = 14),
        strip.background = element_blank()
        )) 
colors <-paletteer_d("ggthemes::Miller_Stone")
```

1. Load in data
```{r}
library(tidyverse)

# Folder containing all CSV files
pilot.data.folder <- "~/GitHub/sporeAI/data/infection.pilot.results"

# Get all CSV files
files <- list.files(
  path = pilot.data.folder,
  pattern = "\\.csv$",
  full.names = TRUE
)
# Since folder contains .csvs
all_data <- map_dfr(files, function(file) {

  filename <- tools::file_path_sans_ext(basename(file))

  dat <- read_csv(file, show_col_types = FALSE)

  parts <- str_split(filename, "_")[[1]]

  dat %>%
    mutate(
      Treatment = parts[1],
      Time = as.numeric(str_remove(parts[2], "T")),
      Stage = if_else(
        "preinfect" %in% parts,
        "preinfect",
        "postinfect"
      ),
      Replicate = as.numeric(parts[length(parts)]),
      File = filename
    )
})
```

2. Rename treatments to be more informative
```{r}
all_data <- all_data %>%
  mutate(
    Flask = Treatment,
    Treatment = recode(
      Treatment,
      "A" = "Early infection",
      "B" = "Early infection",
      "C" = "Late infection",
      "D" = "Control"
    ),
    BioRep = recode(
      Flask,
      "A" = "1",
      "B" = "2",
      "C" = "1",
      "D" = "1"
    )
  )

# Rename replicate to image
all_data <- all_data %>%
  rename(
    Image = Replicate)
head(all_data)
```

2. Looking at the distributions
```{r}
glimpse(all_data)
summary(all_data)

all_data %>%
  count(name, sort = TRUE)

# Check the amount of data per experimental group
all_data %>%
  count(Treatment, Flask, Time, Image)

filt.data <- all_data %>%
  filter(Stage != "preinfect") # subset from the preinfect images 
```
```{r}
counts <- filt.data %>%
  count(Treatment, Flask, Time, Image, name)

ggplot(
  counts,
  aes(
    x = Time,
    y = n,
    color = name,
    group = interaction(Flask, Image, name)
  )
) +
  geom_line(alpha = 0.4) +
  geom_point() +
  facet_wrap(~ Treatment) +
  labs(
    x = "Time",
    y = "Number of cells",
    color = "Stage"
  )
```
## Summarize as percentage
```{r}
image_summary <- filt.data %>%
  count(Treatment, Flask, Time, Image, name) %>%
  group_by(Treatment, Flask, Time, Image) %>%
  mutate(
    total = sum(n),
    percent = n / total * 100
  ) %>%
  ungroup()
```

## Plot as percent of cells
```{r}
ggplot(
  image_summary,
  aes(
    x = Time,
    y = percent,
    color = name
  )
) +
  geom_point() +
  geom_line(
    aes(group = interaction(Flask, Image, name)),
    alpha = 0.3
  ) +
  facet_wrap(~ Treatment) +
  labs(
    x = "Time",
    y = "Percent of cells",
    color = "Stage"
  )
```

## Plot early infection
```{r}
image_summary <- all_data %>%
  count(Treatment, Flask, Time, Image, name) %>%
  group_by(Treatment, Flask, Time, Image) %>%
  mutate(
    total = sum(n),
    percent = n / total * 100
  ) %>%
  ungroup()
ggplot(
  image_summary %>%
    filter(Treatment == "Early infection"),
  aes(
    x = Time,
    y = percent,
    color = name
  )
) +
  geom_point() +
  geom_line(
    aes(group = interaction(Flask, Image, name)),
    alpha = 0.3
  ) +
  labs(
    title = "Early infection",
    x = "Time",
    y = "Percent of cells",
    color = "Cell type"
  ) 

```
## Plot late infection
```{r}
ggplot(
  image_summary %>%
    filter(Treatment == "Late infection"),
  aes(
    x = Time,
    y = percent,
    color = name
  )
) +
  geom_point() +
  geom_line(
    aes(group = interaction(Flask, Image, name)),
    alpha = 0.3
  ) +
  labs(
    title = "Late infection",
    x = "Time",
    y = "Percent of cells",
    color = "Cell type"
  )
```
## Control
```{r}
ggplot(
  image_summary %>%
    filter(Treatment == "Control"),
  aes(
    x = Time,
    y = percent,
    color = name
  )
) +
  geom_point() +
  geom_line(
    aes(group = interaction(Flask, Image, name)),
    alpha = 0.3
  ) +
  labs(
    title = "Control",
    x = "Time",
    y = "Percent of cells",
    color = "Cell type"
  )
```
```{r}
# Look at flasks A and B separately (they are replicates)
ggplot(
  image_summary %>%
    filter(Flask %in% c("A", "B")),
  aes(
    x = Time,
    y = percent,
    color = name
  )
) +
  geom_point() +
  geom_line(
    aes(group = interaction(Image, name)),
    alpha = 0.4
  ) +
  facet_wrap(~ Flask) +
  labs(
    x = "Time",
    y = "Percent of cells",
    color = "Cell type"
  ) 
```
## Summarize data to mean percent of cells for each stage and plot over time for each treatment (accounts for flask A and B in early infection)
```{r}
image_summary <- filt.data %>%
  count(Treatment, Flask, Time, Image, name) %>%
  group_by(Treatment, Flask, Time, Image) %>%
  mutate(
    total = sum(n),
    percent = n / total * 100
  ) %>%
  ungroup()


flask_summary <- image_summary %>%
  group_by(Treatment, Flask, Time, name) %>%
  summarise(
    mean_percent = mean(percent, na.rm = TRUE),
    sd_percent = sd(percent, na.rm = TRUE),
    n_images = n(),
    se_percent = sd_percent / sqrt(n_images),
    .groups = "drop"
  )

ggplot(
  flask_summary,
  aes(
    x = Time,
    y = mean_percent,
    color = name,
    group = name
  )
) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  geom_errorbar(
    aes(
      ymin = mean_percent - se_percent,
      ymax = mean_percent + se_percent
    ),
    width = 0.15
  ) +
  facet_wrap(~Treatment, nrow=1) +
  labs(
    x = "Time",
    y = "Mean percent of cells",
    color = "Cell type"
  ) 

plot1 <- ggplot(
  flask_summary,
  aes(
    x = Time,
    y = mean_percent,
    color = name,
    group = name
  )
) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  geom_errorbar(
    aes(
      ymin = mean_percent - se_percent,
      ymax = mean_percent + se_percent
    ),
    width = 0.15
  ) +
  facet_wrap(~ Treatment, nrow = 1) +
  labs(
    x = "Time",
    y = "Mean percent of cells",
    color = "Cell type"
  )

plot1

ggsave(
  "treatments.png",
  plot = plot1,
  width = 15,
  height = 5,
  units = "in",
  dpi = 300
)
```

# Plot with the lines but also smooth to get sense of what is happening
```{r}
plot2 <- ggplot(
  flask_summary, # plotting this does not account for the fact that flask a and b are replicates, so I plotted differently below
  aes(
    x = Time,
    y = mean_percent,
    color = name,
    group = name
  )
) +
  geom_line(
    alpha = 0.25,
    linewidth = 0.6
  ) +
  geom_point(
    alpha = 0.5,
    size = 2
  ) +
  geom_smooth(
    method = "loess",
    se = FALSE,
    linewidth = 1.2
  ) +
  facet_wrap(~ Treatment, nrow = 1) +
  labs(
    x = "Time",
    y = "Mean percent of cells",
    color = "Cell type"
  )

plot2

ggsave(
  "treatments_smooth.png",
  plot = plot2,
  width = 15,
  height = 5,
  units = "in",
  dpi = 300
)
```

# look at distributions of cells via stacked bar
```{r}
stage_summary <- filt.data %>%
  count(Treatment, Time, name) %>%
  group_by(Treatment, Time) %>%
  mutate(
    percent = n / sum(n) * 100
  ) %>%
  ungroup()

ggplot(
  stage_summary,
  aes(
    x = factor(Time),
    y = percent,
    fill = name
  )
) +
  geom_col() +
  facet_wrap(~ Treatment, nrow = 1) +
  labs(
    x = "Time",
    y = "Percent of cells",
    fill = "Sporulation stage"
  ) 
```

## Diag plot to see model confidence in each stage
```{r}
plot3 <- ggplot(
  filt.data,
  aes(
    x = name,
    y = confidence
  )
) +
  geom_boxplot() +
  labs(
    x = "Predicted sporulation stage",
    y = "Model confidence"
  )

plot3
# ggsave(
#   "model_confidence.png",
#   plot = plot3,
#   width = 15,
#   height = 5,
#   units = "in",
#   dpi = 300
# )
```
## 
```{r}
# average biological replicates (correcting for A and B in early infection treatment)
treatment_summary <- flask_summary %>%
  group_by(Treatment, Time, name) %>%
  summarise(
    mean_percent = mean(mean_percent, na.rm = TRUE),
    .groups = "drop"
  )

plot4 <- ggplot(
  treatment_summary,
  aes(
    x = Time,
    y = mean_percent,
    color = name,
    group = name
  )
) +
  geom_point(size = 2) +
  geom_line(alpha = 0.3) +
  geom_smooth(
    method = "loess",
    se = FALSE,
    linewidth = 1.2
  ) +
  facet_wrap(~ Treatment, nrow = 1) +
  labs(
    x = "Time",
    y = "Mean percent of cells",
    color = "Cell type"
  )
plot4
# ggsave(
#   "treatments_smooth_mean.png",
#   plot = plot4,
#   width = 15,
#   height = 5,
#   units = "in",
#   dpi = 300
# )
```


## Initial thoughts
- This pilot experiment likely does not capture the full sporulation window: Across all experimental flasks, mature spores are very low, I expected to see an increase in mature spores over time across all treatments (especially the control), this indicates to me that I did not capture the full sporulation window (which was my initial gut feeling when looking a the images--I want to continue experiment to image hours 8-16)
- On this same note, even though late stage spores and mature spore abundance is low, it’s the highest (accounts for ~ 35% ) in early infection treatment at t = 5 which is curious
- Early infection shows a pronounced decline in vegetative cells between approx T2-T5, this could be a signal of lysis maybe, and that is interestingly accompanied by an increase in stage 4 spores
- I wasn’t expecting late-infection treatment to show a rather consistent amount of vegetative cells over time, I think this could be due to the model not yet trained on stage 1 of sporulation (since veg. cells and stage 1 cells are phenotyptically similar), however if this persists it could be indicative of some sort of stalling of the progression of sporulation

## Plotting with A and B as two separate lines in early infection treatment 
```{r}
ggplot(
  flask_summary,
  aes(
    x = Time,
    y = mean_percent,
    color = name
  )
) +
  geom_line(
    aes(group = interaction(Flask, name)),
    alpha = 0.25,
    linewidth = 0.6
  ) +
  geom_point(
    alpha = 0.5,
    size = 2
  ) +
  geom_smooth(
    aes(group = name),
    method = "loess",
    se = FALSE,
    linewidth = 1.2
  ) +
  facet_wrap(~ Treatment, nrow = 1) +
  labs(
    x = "Time",
    y = "Mean percent of cells",
    color = "Cell type"
  )
```

